---
title: "Lab sessions 5 - Bayesian Statistics"
date: "May, 6 2021"
output:
  html_document:
    toc: yes
  beamer_presentation:
    highlight: tango
  include: null
  ioslides_presentation:
    highlight: tango
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
  slide_level: 2
  slidy_presentation:
    fig.height: 3
  fig.width: 4
  highlight: tango
header-includes:
- \usepackage{color}
- \usepackage{tcolorbox}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
graphics: yes
subtitle: 
fontsize: 10pt
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.625, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/',  eval=TRUE, cache=TRUE)
library(MASS)
```
```{r setup, include=FALSE}
library(knitr)
local({
    hook_plot = knit_hooks$get('plot')
    knit_hooks$set(plot = function(x, options) {
        paste0('\n\n----\n\n', hook_plot(x, options))
    })
})
```

# Cockroaches

## Recap
We are analysing data about complaints for the presence of cocroaches in 10 residential buildings. Using a Bayesian framework, we modelled the number of complaints using a Poisson model. Diagnostic analysis of the model highlights some residual variability. We decided to try to improve our estimates adding one predictor and an exposure term to the Poisson model, but the situation did not improve. The 'Poisson' assumption on the model variance turned out to be too restrictive to model our data. The Negative Binomial model gave better results, but examining the standardized residual plot we noticed that some of them were still very large.

A posterior predictive check considering that the data are clustered by building.

```{r,eval=TRUE,echo=TRUE, warning=FALSE, results='hide',message=FALSE}
library(rstan)
library(dplyr)
library(lubridate)
library(ggplot2)
library(bayesplot)

theme_set(bayesplot::theme_default())
rstan_options(auto_write=TRUE)
```
```{r,eval=TRUE,echo=TRUE}
pest_data <- readRDS('pest_data.RDS')
str(pest_data)
summary(pest_data)

##number of buildings
N_buildings <- length(unique(pest_data$building_id))
N_buildings
```

```{r stan-data}
## arrange data into a list
stan_dat_simple <- list(
  N = nrow(pest_data), 
  complaints = pest_data$complaints,
  traps = pest_data$traps,
  log_sq_foot = log(pest_data$total_sq_foot/1e4),
  live_in_super = pest_data$live_in_super
)
str(stan_dat_simple)
```

```{r , cache=TRUE, results="hide", message=FALSE}
comp_model_NB <- stan_model('multiple_NB_regression.stan')
```

```{r runNB}
fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_simple)
samps_NB <- rstan::extract(fitted_model_NB)
```

```{r ppc-group_means}
y_rep <- samps_NB$y_rep

ppc_stat_grouped(
  y = stan_dat_simple$complaints, 
  yrep = y_rep, 
  group = pest_data$building_id, 
  stat = 'mean',
  binwidth = 0.2
)
```

## Lab 5

Select some of the variables containing information at the building level (live in super, age of building, average tentant age, and monthly average rent) and arrange them into a matrix called `building_data` (nrow = number of buildings).

```{r prep-data}
N_months <- length(unique(pest_data$date))

# Add some IDs for building and month
pest_data <- pest_data %>%
  mutate(
    building_fac = factor(building_id, levels = unique(building_id)),
    building_idx = as.integer(building_fac),
    ids = rep(1:N_months, N_buildings),
    mo_idx = lubridate::month(date)
  )

# Center and rescale the building specific data
building_data <- pest_data %>%
    select(
      building_idx,
      live_in_super,
      age_of_building,
      average_tenant_age,
      monthly_average_rent
    ) %>%
    unique() %>%
    arrange(building_idx) %>%
    select(-building_idx) %>%
    scale(scale=FALSE) %>%
    as.data.frame() %>%
    mutate( # scale by constants
      age_of_building = age_of_building / 10,
      average_tenant_age = average_tenant_age / 10,
      monthly_average_rent = monthly_average_rent / 1000
    ) %>%
    as.matrix()

str(building_data)

stan_dat_hier <-
  with(pest_data,
        list(complaints = complaints,
             traps = traps,
             N = length(traps),
             J = N_buildings,
             log_sq_foot = log(pest_data$total_sq_foot/1e4),
             building_data = building_data,
             mo_idx = as.integer(as.factor(date)),
             K = 4,
             building_idx = building_idx
             )
       )
str(stan_dat_hier)
```

Model the variation across buildings using a varying intercept Negative Binomial hierarchical model.

Using stan write, compile and fit the following varying intercept model

$$
\text{complaints}_{ib} \sim \text{Neg-Binomial}(\lambda_{ib}, \phi) \\
\lambda_{ib}  = \exp{(\eta_{ib})} \\
\eta_{ib} = \mu_{b(i)} + \beta \, {\rm traps}_{i} + \text{log_sq_foot}_i \\
\mu_b \sim \text{Normal}(\alpha + \beta_{\rm super}\, {\rm super}_b + \ldots + \beta_{\rm mar}\, {\rm mar}_b , \sigma_{\mu})
$$

```{r comp-NB-hier, cache=TRUE, results="hide", message=FALSE}
comp_model_NB_hier <- stan_model('hier_NB_regression.stan')
```
```{r run-NB-hier}
fitted_model_NB_hier <- sampling(comp_model_NB_hier, data = stan_dat_hier,
  chains = 4, #cores = 4, 
  iter = 4000)

```
Do you have any divergent transitions?

```{r}
samps_hier_NB <- rstan::extract(fitted_model_NB_hier)
```

```{r print-NB-hier}
print(fitted_model_NB_hier, pars = c('sigma_mu','beta','alpha','phi','mu'))
```
```{r}
mcmc_trace(
  as.array(fitted_model_NB_hier,pars = 'sigma_mu'),
  np = nuts_params(fitted_model_NB_hier),
  window = c(500,1000)
)
```
```{r}
# assign to object so we can compare to another plot later
scatter_with_divs <- mcmc_scatter(
  as.array(fitted_model_NB_hier),
  pars = c("mu[4]", 'sigma_mu'),
  transform = list('sigma_mu' = "log"),
  np = nuts_params(fitted_model_NB_hier)
)
scatter_with_divs
```


```{r}
N_sims <- 1000
log_sigma <- rep(NA, N_sims)
theta <- rep(NA, N_sims)
for (j in 1:N_sims) {
  log_sigma[j] <- rnorm(1, mean = 0, sd = 1)
  theta[j] <- rnorm(1, mean = 0, sd = exp(log_sigma[j]))
}
draws <- cbind("mu" = theta, "log(sigma_mu)" = log_sigma)
mcmc_scatter(draws)
```


```{r}
parcoord_with_divs <- mcmc_parcoord(
  as.array(fitted_model_NB_hier, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier)
)
parcoord_with_divs
```

 We see evidence that our problems concentrate when $\texttt{sigma_mu}$ is small. Reparametrize the model using the non-centered parametrization and check diagnostics.
The distribution of $\texttt{mu}$ does not change:
$$
\mu_b \sim \text{Normal}(\alpha + \texttt{building_data} \, \zeta, \sigma_{\mu})
$$
We add an auxiliary variable in the parameters block $\texttt{mu_raw}\sim \text{Normal}(0, 1)$ and we move $\texttt{mu}$ in the transformed parameters block.

```
transformed parameters {
  vector[J] mu;
  mu = alpha + building_data * zeta + sigma_mu * mu_raw;
}
```


```{r comp-NB-hier-ncp, cache=TRUE}
comp_model_NB_hier_ncp <- stan_model('hier_NB_regression_ncp.stan')
```

```{r run-NB-hier-ncp}
fitted_model_NB_hier_ncp <- sampling(comp_model_NB_hier_ncp, data = stan_dat_hier, #cores=4,
                                     chains = 4)
```

```{r n-eff-NB-hier-ncp-check}
print(fitted_model_NB_hier_ncp, pars = c('sigma_mu','beta','alpha','phi','mu'))
```

```{r}
scatter_no_divs <- mcmc_scatter(
  as.array(fitted_model_NB_hier_ncp),
  pars = c("mu[4]", 'sigma_mu'),
  transform = list('sigma_mu' = "log"),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
bayesplot_grid(scatter_with_divs, scatter_no_divs,
               grid_args = list(ncol = 2), ylim = c(-11, 1))
```

```{r}
parcoord_no_divs <- mcmc_parcoord(
  as.array(fitted_model_NB_hier_ncp, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
bayesplot_grid(parcoord_with_divs, parcoord_no_divs,
               ylim = c(-3, 3))
```

```{r samps-full-hier}
samps_NB_hier_ncp <- rstan::extract(fitted_model_NB_hier_ncp, pars = c('y_rep','inv_phi'))
```


```{r ppc-full-hier}
y_rep <- as.matrix(fitted_model_NB_hier_ncp, pars = "y_rep")
ppc_dens_overlay(stan_dat_hier$complaints, y_rep[1:200,])
```


```{r ppc-group_means-hier}
ppc_stat_grouped(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  group = pest_data$building_id,
  stat = 'mean',
  binwidth = 0.5
)
```


```{r}
ppc_stat_grouped(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  group = pest_data$building_id,
  stat = 'sd',
  binwidth = 0.5
)
```

```{r}
ppc_intervals(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  x = stan_dat_hier$traps
) +
  labs(x = "Number of traps", y = "Number of complaints")
```
```{r}
mean_y_rep <- colMeans(y_rep)
mean_inv_phi <- mean(as.matrix(fitted_model_NB_hier_ncp, pars = "inv_phi"))
std_resid <- (stan_dat_hier$complaints - mean_y_rep) / sqrt(mean_y_rep + mean_y_rep^2*mean_inv_phi)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```

Try to expand the previous model with varying slope.

$$
\text{complaints}_{ib} \sim \text{Neg-Binomial}(\lambda_{ib}, \phi)  \\
\lambda_{ib} = \exp{(\eta_{ib})}\\
\eta_{ib} = \mu_{b(i)} + \kappa_{b(i)} \, \texttt{traps}_{ib} + \text{log_sq_foot}_i \\
\mu_b \sim \text{Normal}(\alpha + \texttt{building_data} \, \zeta, \sigma_{\mu}) \\
\kappa_b \sim \text{Normal}(\beta + \texttt{building_data} \, \gamma, \sigma_{\kappa})
$$

```{r}
stan_dat_hier <- readRDS('pest_data_longer_stan_dat.RDS')
```

```{r comp-NB-hier-slopes, cache=TRUE, results="hide", message=FALSE}
comp_model_NB_hier_slopes <- stan_model('hier_NB_regression_ncp_slopes_mod.stan')
```

```{r run-NB-hier-slopes}
fitted_model_NB_hier_slopes <-
  sampling(
    comp_model_NB_hier_slopes,
    data = stan_dat_hier,
    chains = 4, #cores = 4,
    control = list(adapt_delta = 0.95)
  )
```

```{r}
mcmc_hist(
  as.matrix(fitted_model_NB_hier_slopes, pars = "sigma_kappa"),
  binwidth = 0.005
)
```

```{r}
print(fitted_model_NB_hier_slopes, pars = c('kappa','beta','alpha','phi','sigma_mu','sigma_kappa','mu'))
```

```{r}
mcmc_hist(
  as.matrix(fitted_model_NB_hier_slopes, pars = "beta"),
  binwidth = 0.005
)
```

```{r ppc-full-hier-slopes}
y_rep <- as.matrix(fitted_model_NB_hier_slopes, pars = "y_rep")
ppc_dens_overlay(
  y = stan_dat_hier$complaints,
  yrep = y_rep[1:200,]
)
```






