---
title: "Lab sessions 4 - Bayesian Statistics"
date: "April, 26 2021"
output:
  html_document:
    toc: yes
  beamer_presentation:
    highlight: tango
  include: null
  ioslides_presentation:
    highlight: tango
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
  slide_level: 2
  slidy_presentation:
    fig.height: 3
  fig.width: 4
  highlight: tango
header-includes:
- \usepackage{color}
- \usepackage{tcolorbox}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
institute: University of Udine & University of Trieste
graphics: yes
subtitle: 
fontsize: 10pt
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE,  dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/',  eval=TRUE, cache=TRUE)
library(MASS)
```
```{r setup, include=FALSE}
library(knitr)
local({
    hook_plot = knit_hooks$get('plot')
    knit_hooks$set(plot = function(x, options) {
        paste0('\n\n----\n\n', hook_plot(x, options))
    })
})
```


## Cockroaches

**1** Load and explore the dataset in the file `pest_data.RDS`. The dataset includes 14 variables, that are:

* `complaints`: Number of complaints per building per month
* `building_id`: The unique building identifier
* `traps`: The number of traps used per month per building
* `date`: The date at which the number of complaints are recorded
* `live_in_super`: An indicator for whether the building has a live-in super
* `age_of_building`: The age of the building
* `total_sq_foot`: The total square footage of the building
* `average_tenant_age`: The average age of the tenants per building
* `monthly_average_rent`: The average monthly rent per building
* `floors`: The number of floors per building

<!-- ```{r,eval=TRUE,echo=FALSE} -->
<!-- knitr::opts_chunk$set( -->
<!--   echo = TRUE,  -->
<!--   dev = "png", -->
<!--   dpi = 150, -->
<!--   fig.align = "center", -->
<!--   comment = NA -->
<!--   ) -->
<!-- ``` -->
```{r,eval=TRUE,echo=TRUE, warning=FALSE, results='hide',message=FALSE}
library(rstan)
library(dplyr)
library(lubridate)
library(ggplot2)
library(bayesplot)

theme_set(bayesplot::theme_default())

# seed for R's pseudo-RNGs, not Stan's
set.seed(1123) 
```

```{r,eval=TRUE,echo=TRUE}
pest_data <- readRDS('pest_data.RDS')
str(pest_data)
summary(pest_data)

##number of buildings
N_buildings <- length(unique(pest_data$building_id))
N_buildings
```

```{r,eval=TRUE,echo=FALSE}
ggplot(pest_data, aes(x = complaints)) + 
  geom_bar()

ggplot(pest_data, aes(x = traps, y = complaints, color = live_in_super == TRUE)) + 
    geom_jitter() +
    labs(color='Live-in super')
```
```{r, eval=TRUE,echo=FALSE, data-plots-ts, fig.height = 16,fig.width = 6, fig.height = 8}
ggplot(pest_data, aes(x = date, y = complaints, color = live_in_super == TRUE)) + 
  geom_line(aes(linetype = "Number of complaints")) + 
  geom_point(color = "black") + 
  geom_line(aes(y = traps, linetype = "Number of traps"), color = "black", size = 0.25) + 
  facet_wrap(~building_id, scales = "free", ncol = 2, labeller = label_both) + 
  scale_x_date(name = "Month", date_labels = "%b") + 
  scale_y_continuous(name = "", limits = range(pest_data$complaints)) + 
  scale_linetype_discrete(name = "") + 
  scale_color_discrete(name = "Live-in super")
```

<!-- Knowing that the number of complaints over a month is unlikely to be zero and that rarerly there are a large number of complaints over a month, how would you model the outcome variable `complaints`? -->

**2** Write the following Poisson model in Stan, for $i=1,\ldots,120$

\begin{align*}
	\textrm{complaints}_i & \sim \textrm{Poisson}(\lambda_i), \\
	\lambda_i & = \exp{(\eta_i)} \\
	\eta_i &= \alpha + \beta \, \textrm{traps}_i
\end{align*}
Is the number of complaints associated with the number of traps? 
```{r}
## compile the model
comp_model_P <- stan_model('simple_poisson_regression.stan')
```
```{r stan-data}
## arrange data into a list
stan_dat_simple <- list(
  N = nrow(pest_data), 
  complaints = pest_data$complaints,
  traps = pest_data$traps
)
str(stan_dat_simple)
```
```{r fit_P_real_data, cache=TRUE}
## fit the model
fit_P_real_data <- sampling(comp_model_P, data = stan_dat_simple)
```

```{r results_simple_P}
## print the parameters 
print(fit_P_real_data, pars = c('alpha','beta'))
```

```{r hist_simple_P}
mcmc_hist(as.matrix(fit_P_real_data, pars = c('alpha','beta')))
mcmc_scatter(as.matrix(fit_P_real_data, pars = c('alpha','beta')), alpha = 0.2)
```
It appears that there is association between the number of bait stations and the number of complaints. 

**3** Verify that  the 'Poisson' assumption in the Poisson generalised linear model is not satisfied. Improve your model including the indicator variable for whether the building has a live-in super and the total square footage of the building (offset).

```{r}
## posterior predictive checking
y_rep <- as.matrix(fit_P_real_data, pars = "y_rep")
ppc_dens_overlay(y = stan_dat_simple$complaints, y_rep[1:200,])
```
```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```


```{r}
## standardised residuals of the observed vs predicted number of complaints
mean_y_rep <- colMeans(y_rep)
std_resid <- (stan_dat_simple$complaints - mean_y_rep) / sqrt(mean_y_rep)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```
Residuals are mostly positive, thus the model tends to underestimate the number of complaints.

We expand the model adding the live-in super variable and the total square footage of the building (offset). You can interpret the parameters as rate of complaints per square foot.

```{r}
ggplot(pest_data, aes(x = log(total_sq_foot), y = log1p(complaints))) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

\begin{align*}
	\textrm{complaints}_i & \sim \textrm{Poisson}(\lambda_i) \\
	\lambda_i & = \exp{(\eta_i)} \\
	\eta_i &= \alpha + \beta \, \textrm{traps}_i + \beta_{\rm super} \, {\rm super}_{i} + \text{log_sq_foot}_{i}
\end{align*}

```{r}
## add the two variables to the list of the data
stan_dat_simple$log_sq_foot <- log(pest_data$total_sq_foot/1e4)
stan_dat_simple$live_in_super <- pest_data$live_in_super
str(stan_dat_simple)
```

```{r compmultPDGP}
## compile the model
comp_model_P_mult <- stan_model('multiple_poisson_regression.stan')
```

```{r fit_mult_P_real_dat}
fit_model_P_mult_real <- sampling(comp_model_P_mult, data = stan_dat_simple)
print(fit_model_P_mult_real, pars = c('alpha','beta','beta_super'))
y_rep <- as.matrix(fit_model_P_mult_real, pars = "y_rep")
ppc_dens_overlay(stan_dat_simple$complaints, y_rep[1:200,])
```
```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

We've increased the tails a bit more at the larger numbers of traps but we still have some large observed numbers of complaints that the model would consider extremely unlikely events. 


**4** Write the following model in stan and compare results with the previous model.

\begin{align*}
\text{complaints}_i & \sim \text{Neg-Binomial}(\lambda_i, \phi) \\
\lambda_i & = \exp{(\eta_i)} \\
\eta_i &= \alpha + \beta \, {\rm traps}_i + \beta_{\rm super} \, {\rm super}_{i} + \text{log_sq_foot}_{i}
\end{align*}


In Stan the negative binomial mass function we will use is called 
$\texttt{neg_binomial_2_log}( \text{reals} \, \eta, \text{reals} \, \phi)$. Like the `poisson_log` function, this negative binomial mass function is parameterized in terms of its log-mean, $\eta$, but it also has a precision $\phi$ such that

$$
\mathbb{E}[y] \, = \lambda = \exp(\eta)
$$

$$
\text{Var}[y] = \lambda + \lambda^2/\phi = \exp(\eta) + \exp(\eta)^2 / \phi.
$$

As $\phi$ gets larger, the term $\lambda^2 / \phi$ approaches zero and so the variance of the negative-binomial approaches $\lambda$, i.e., the negative-binomial gets closer and closer to the Poisson.

```{r message=FALSE, cache=TRUE, results="hide"}
comp_model_NB <- stan_model('multiple_NB_regression.stan')
```

```{r runNB}
fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_simple)
#fitted_model_NB <- stan(file="multiple_NB_regression.stan", data=stan_dat_simple)
samps_NB <- rstan::extract(fitted_model_NB)
```

```{r ppc-full}
## predictions vs. the data
y_rep <- samps_NB$y_rep
ppc_dens_overlay(stan_dat_simple$complaints, y_rep[1:200,])
```

```{r}
## standardised residuals
mean_inv_phi <- mean(samps_NB$inv_phi)
mean_y_rep <- colMeans(y_rep)
std_resid <- (stan_dat_simple$complaints - mean_y_rep) / sqrt(mean_y_rep + mean_y_rep^2*mean_inv_phi)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```

```{r}
## prediction by number of traps
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

Looks OK but... Data are clustered by building. A posterior predictive check can help us understanding if it would be a good idea to add the building information into the model.

```{r ppc-group_means}
ppc_stat_grouped(
  y = stan_dat_simple$complaints, 
  yrep = y_rep, 
  group = pest_data$building_id, 
  stat = 'mean',
  binwidth = 0.2
)
```

We're getting plausible predictions for most building means but some are
estimated better than others and some have larger uncertainties than we might
expect. If we explicitly model the variation across buildings we may be able to
get much better estimates.

... to be continued

